---
title: "Cats Cradle Spatial Vignette"
output: html_notebook
---

If the xenium object is downloaded we can read this to enable the 
spatial visualisation of results. 
```{r}
#xenium.obj = read.rds("/path/to/xenium.obj.rds")
```

Compute spatial neighbours based on Delaunay triangulation.
```{r}
delaunayNeighbours = computeNeighboursDelaunay(centroids)
```

Alternatively compute neighbours based on Euclidean distance.
```{r}
euclideanNeighbours = computeNeighboursEuclidean(centroids, 20)
```

Compute neighbourhoods, here each a neighbourhood is computed for each cell that 
comprises of its 1st degree neighbours. Set "addSelf = T" to include the central
cell in its neighbourhood.
```{r}
neighbourhood = computeNeighbourhoods(delaunayNeighbours, names(clusters))
```

Compute a matrix where neighbourhoods are rows and cell types are columns. 
The values in the matrix indicate the number of cells of a given type within a 
neighbourhood.
```{r}
neighbourhoodbyCTmatrix = computeNeighbourhoodByCTMatrix(neighbourhood, 
                                                         clusters)
```

Create a seurat object using a neighbourhood by cell type matrix and calculate 
UMAP and clusters. As the dimensionlity of the feature space is relatively low 
(number of cell types) we calculate the UMAP using features rather than PCs.

```{r}
neighbourhoodSeurat = computeNeighbourhoodByCTSeurat(neighbourhoodbyCTmatrix)
```

As there are many clusters (46) we can also cluster at a lower resolution to 
achieve a less granular clustering.
```{r}
neighbourhoodSeurat = FindClusters(neighbourhoodSeurat,resolution=0.01)
```
Add cell type information to the neighbourhoodSeurat object.
```{r}
neighbourhoodSeurat$cellType = as.character(clusters)
```

Visualise neighbourhood clusters.
```{r}
DimPlot(neighbourhoodSeurat, group.by = c("cellType", "seurat_clusters"), cols = "polychrome", reduction = "umap")
```


Create seurat object for a cell type by neighbourhood matrix. Here we have a 
large feature space (numbers of cells) however low number of observations 
(number of cell types). Therefore we compute the UMAP using PCs however due to 
the low number of observations we need to set a lower value for n.neighbours.
```{r}
neighbourhoodSeuratPrime = 
  computeNeighbourhoodByCTSeurat(t(neighbourhoodbyCTmatrix), npcs = 20, transpose = T, resolution = 1, n.neighbors = 5)

neighbourhoodSeuratPrime$cellType = colnames(neighbourhoodSeuratPrime)

DimPlot(neighbourhoodSeuratPrime, group.by = "cellType", cols = "polychrome", reduction = "umap", label = T)
```

We can also compute a graph embedding.
```{r}
neighbourhoodSeuratPrime = computeGraphEmbedding(neighbourhoodSeuratPrime)

DimPlot(neighbourhoodSeuratPrime,group.by = "cellType", cols = "alphabet", reduction = "graph", label = T)
```

We achieve only one cluster using the louvain algorithm. Let's try hierachical 
clustering.
```{r}
library(pheatmap)

pca = Embeddings(neighbourhoodSeuratPrime, reduction = "pca")
res = pheatmap(pca)
```

Here we decide to cut the tree at the correct level to form 3 clusters.
```{r}
CTclust = cutree(res$tree_row, k = 3)
neighbourhoodSeuratPrime$seurat_clusters = factor(CTclust)
```

Lets investigate how neighbourhood clusters relate to cell type clusters.
```{r}

averageExpMatrix = getAverageExpressionMatrix(neighbourhoodSeurat,neighbourhoodSeuratPrime)
averageExpMatrix = tagRowAndColNames(averageExpMatrix,
                                     ccTag='neighbourhoodClusters_',
                                     gcTag='cellTypeClusters_')
pheatmap(averageExpMatrix,
      treeheight_row=0,
      treeheight_col=0,
      fontsize_row=8,
      fontsize_col=8,
      cellheight=10,
      cellwidth=10)


sankeyFromMatrix(averageExpMatrix)
```
We can also investigate how cell clusters based on their neighbourhoods (here we call this domain) relate 
to cell clusters based on gene expression (here we call this CT - cell type).

```{r}
CTDomain = table(neighbourhoodSeurat$cellType,neighbourhoodSeurat@active.ident)
CTDomain = CTDomain/rowSums(CTDomain)

pheatmap(CTDomain[,],
      treeheight_row=0,
      treeheight_col=0,
      fontsize_row=8,
      fontsize_col=8,
      cellheight=10,
      cellwidth=10)

sankeyFromMatrix(CTDomain)
```
If we have the xenium object loaded we can visualise cell types and 
neighbourhood clusters spatially.
```{r}
ImageDimPlot(xenium.obj, cols = "polychrome", size = 0.75)
```
```{r}
xenium.obj$neighbourhoodCluster = factor(as.character(neighbourhoodSeurat$RNA_snn_res.0.01)) 
ImageDimPlot(xenium.obj, cols = "polychrome", size = 0.75, 
             group.by = "neighbourhoodCluster")
```

